---
title: "LLM data import and cleaning"
format: html
---

::: {.Rcode title="Source code"}

The full R script is available at:

- [`R/Analysis data import and cleaning.R`](https://github.com/VagishHemmige/USRDS-PDF-analysis-project/blob/master/R/Analysis%20data%20import%20and%20cleaning.R)

This R script relies on the following helper files:

- [`R/Analysis setup.R`](https://github.com/VagishHemmige/USRDS-PDF-analysis-project/blob/master/R/Analysis%20setup.R)
- [`R/functions.R`](https://github.com/VagishHemmige/USRDS-PDF-analysis-project/blob/master/R/functions.R)
:::



## Import raw data

The analysis begins by importing the raw outputs from each large language model together with the human-reviewed abstraction file. These inputs were produced in different formats, so they must first be standardized before they can be compared directly.

```{r, eval=FALSE}
final_results_ChatGPT_raw <- readRDS("Results/ChatGPT/final_results_ChatGPT.rds")

summary_final_claude_raw <- readRDS("Results/Claude/summary_final_claude.rds") %>%
  bind_rows()

summary_final_gemini_raw <- read_csv("Results/Gemini/summary_output_gemini.csv")

human_results_raw <- read_excel(
  "data-raw/Data abstraction Excel reviewer 1 and 2.xlsx",
  sheet = "Reviewer 2"
) %>%
  clean_names()
```

## Clean model outputs

Each model output is then cleaned separately. A model-specific suffix is added to each variable name so that the human-reviewed and model-generated versions of the same field can coexist in a single merged dataset. Unneeded text-heavy fields are removed from the ChatGPT output to keep the working dataset focused on structured variables.

```{r, eval=FALSE}
final_results_ChatGPT_clean <- final_results_ChatGPT_raw %>%
  rename_with(~ paste0(.x, "_ChatGPT"), -filename) %>%
  select(-text_ChatGPT, -raw_json_ChatGPT)

summary_final_claude_clean <- summary_final_claude_raw %>%
  rename_with(~ paste0(.x, "_claude"), -filename)

summary_final_gemini_clean <- summary_final_gemini_raw %>%
  rename_with(~ paste0(.x, "_gemini"), -filename)
```

## Derive structured variables from the human review file

The human review spreadsheet requires additional preprocessing. Several variables were originally captured as free-text fields listing programming languages, USRDS files, or abstracted data components. These were split into separate binary indicator variables so that they could be compared directly with the corresponding structured outputs from the language models.

```{r, eval=FALSE}
human_results_clean <- human_results_raw %>%
  mutate(languages_used_R = str_detect(what_programming_language_s_was_were_used, "R")) %>%
  mutate(languages_used_SAS = str_detect(what_programming_language_s_was_were_used, "SAS")) %>%
  mutate(languages_used_Stata = str_detect(what_programming_language_s_was_were_used, "STATA")) %>%
  mutate(languages_used_SPSS = str_detect(what_programming_language_s_was_were_used, "SPSS")) %>%
  mutate(languages_used_Python = str_detect(what_programming_language_s_was_were_used, "Python")) %>%
  mutate(languages_used_SQL = str_detect(what_programming_language_s_was_were_used, "SQL")) %>%
  mutate(languages_used_Julia = str_detect(what_programming_language_s_was_were_used, "Julia")) %>%
  mutate(languages_used_Matlab = str_detect(what_programming_language_s_was_were_used, "Matlab")) %>%
  mutate(files_used_Core = str_detect(what_files_were_used, "Core")) %>%
  mutate(files_used_Institution = str_detect(what_files_were_used, "Institution")) %>%
  mutate(files_used_PhysicianSupplier = str_detect(what_files_were_used, "Physician/Supplier")) %>%
  mutate(files_used_Hospital = str_detect(what_files_were_used, "Hospital")) %>%
  mutate(files_used_Transplant = str_detect(what_files_were_used, "Transplant")) %>%
  mutate(files_used_PartD = str_detect(what_files_were_used, "Part D")) %>%
  mutate(files_used_CROWNWeb = str_detect(what_files_were_used, "CROWNWeb")) %>%
  mutate(component_basicdemographics = str_detect(which_pieces_of_data_were_abstracted_from_the_files, "Basic demographics")) %>%
  mutate(component_icd9 = str_detect(which_pieces_of_data_were_abstracted_from_the_files, "ICD9 code")) %>%
  mutate(component_icd10 = str_detect(which_pieces_of_data_were_abstracted_from_the_files, "ICD10 code")) %>%
  mutate(component_cptcode = str_detect(which_pieces_of_data_were_abstracted_from_the_files, "CPT code")) %>%
  mutate(component_medicationD = str_detect(which_pieces_of_data_were_abstracted_from_the_files, "Medications")) %>%
  mutate(component_transplant = str_detect(which_pieces_of_data_were_abstracted_from_the_files, "Transplant-specific information")) %>%
  mutate(component_costs = str_detect(which_pieces_of_data_were_abstracted_from_the_files, "Costs")) %>%
  mutate(component_residential = str_detect(which_pieces_of_data_were_abstracted_from_the_files, "Residential location")) %>%
  select(
    -timestamp, -changes_made, -comments_16, -comments_26, -confirmed_by_vh,
    -what_files_were_used, -what_programming_language_s_was_were_used,
    -which_crosswalks_were_used
  ) %>%
  rename(filename = filepath) %>%
  rename_with(~ paste0(.x, "_human"), -filename) %>%
  mutate(across(
    where(~ all(.x %in% c("Yes", "No", NA))),
    ~ case_when(
      .x == "Yes" ~ TRUE,
      .x == "No"  ~ FALSE,
      TRUE ~ NA
    )
  )) %>%
  mutate(
    filename = if_else(
      str_ends(filename, "\\.pdf"),
      filename,
      paste0(filename, ".pdf")
    )
  )
```


## Check filename alignment across sources

Before the datasets are merged, filenames are checked to ensure that the same PDF records are present across the human-reviewed and model-generated datasets. These checks make it easier to identify missing or unmatched files before proceeding.

```{r, eval=FALSE}
setdiff(human_results_clean$filename, final_results_ChatGPT_clean$filename)
setdiff(final_results_ChatGPT_clean$filename, human_results_clean$filename)

setdiff(human_results_clean$filename, summary_final_claude_clean$filename)
setdiff(summary_final_claude_clean$filename, human_results_clean$filename)

setdiff(human_results_clean$filename, summary_final_gemini_clean$filename)
setdiff(summary_final_gemini_clean$filename, human_results_clean$filename)

# Gemini unable to parse a single PDF despite multiple attempts
```


## Merge cleaned datasets

Once the inputs have been standardized and checked, they are merged into one wide analysis dataset using the shared `filename` field. This produces a single record per paper containing the human-reviewed variables and the corresponding outputs from ChatGPT, Claude, and Gemini.

```{r, eval=FALSE}
df_wide <- human_results_clean %>%
  left_join(final_results_ChatGPT_clean, by = join_by(filename)) %>%
  left_join(summary_final_claude_clean, by = join_by(filename)) %>%
  left_join(summary_final_gemini_clean, by = join_by(filename)) %>%
  filter(!is.na(filename))
```


## Standardize logical variables

After merging, the dataset is reordered and all human and model variables are converted to logical format. This step is important because the imported data may still contain character encodings such as `"yes"` or `"no"` that need to be normalized before agreement statistics are calculated.

```{r, eval=FALSE}
df_wide <- df_wide %>%
  select(filename, sort(setdiff(names(.), "filename")))

logical_cols <- names(df_wide)[
  grepl("_ChatGPT$|_claude$|_gemini$|_human$", names(df_wide))
]

df_wide <- df_wide %>%
  mutate(across(all_of(logical_cols), convert_to_logical))

# write_xlsx(df_wide, "Results/Cleaned and merged data.xlsx")
```


## Identify human-reviewed analysis variables

Finally, the script identifies the set of human-coded binary variables that will be used in downstream agreement analyses. These variables serve as the reference standard for comparisons with each model.


```{r, eval=FALSE}
human_vars <- names(df_wide)[grepl("_human$", names(df_wide))]
```
