---
title: "Functions"
format: html
---

The code in this script performs a number of useful tasks required for the analysis.

::: {.Rcode title="Source code"}


The full R script is available at:

-   [`R/functions.R`](https://github.com/VagishHemmige/USRDS-PDF-analysis-project/blob/master/R/functions.R)

This R script file is itself reliant on the following helper files:

-   [`R/setup.R`](https://github.com/VagishHemmige/USRDS-PDF-analysis-project/blob/master/R/setup.R)
:::

## Function that safely normalizes to logical

The PDF review dataset contains binary variables that may appear in different formats, including logical values and character encodings such as `"yes"`, `"no"`, `"true"`, `"false"`, `"1"`, and `"0"`. The `convert_to_logical()` function standardizes these inputs into `TRUE`, `FALSE`, or `NA`. This preprocessing step helps ensure that downstream comparisons between human and model-generated labels are performed on consistently formatted data.

```{r, eval=FALSE}

convert_to_logical <- function(x) {
  if (is.logical(x)) return(x)
  
  x_clean <- trimws(tolower(as.character(x)))
  
  case_when(
    x_clean %in% c("true", "yes", "1")  ~ TRUE,
    x_clean %in% c("false", "no", "0") ~ FALSE,
    TRUE ~ NA
  )
}




```

# Compute kappa helper

The `compute_kappa()` function calculates unweighted Cohen’s kappa for a single comparison between a human-reviewed variable and a model-generated variable. It first checks that the model variable is present in the dataset, then removes rows with missing values before calculating kappa. This helper is used when creating agreement summaries across multiple variables and models.

```{r, eval=FALSE}


compute_kappa <- function(hvar, mvar, df_filt) {
  if (!(mvar %in% names(df_filt))) return(NA_real_)
  
  df_k <- df_filt %>%
    select(h = all_of(hvar), m = all_of(mvar)) %>%
    drop_na()
  
  if (nrow(df_k) == 0) return(NA_real_)
  
  kappa2(df_k, weight = "unweighted")$value
}



```

# Function to compute numerators, denominators, percentages

The `get_breakdown()` function creates descriptive summaries for a given human-reviewed variable by pairing it with the corresponding ChatGPT, Claude, and Gemini variables. It counts positive results for each model, calculates denominators and percentages, and formats numerator/denominator strings for use in presentation tables. This function is used to build tables showing how model outputs compare with human abstraction across key study characteristics.

```{r, eval=FALSE}


get_breakdown <- function(hvar, df_filt) {
  
  base <- sub("_human$", "", hvar)
  chat <- paste0(base, "_ChatGPT")
  cla  <- paste0(base, "_claude")
  gem  <- paste0(base, "_gemini")
  
  model_vars <- c(chat, cla, gem)
  model_vars <- model_vars[model_vars %in% names(df_filt)]
  if (length(model_vars) == 0) return(NULL)
  
  df_filt %>%
    select(human = all_of(hvar), all_of(model_vars)) %>%
    mutate(human = as.logical(human)) %>%
    filter(!is.na(human)) %>%
    group_by(human) %>%
    summarise(
      chat_num = if (chat %in% names(.)) sum(!!sym(chat), na.rm = TRUE) else NA_real_,
      cla_num  = if (cla  %in% names(.)) sum(!!sym(cla),  na.rm = TRUE) else NA_real_,
      gem_num  = if (gem  %in% names(.)) sum(!!sym(gem),  na.rm = TRUE) else NA_real_,
      
      den      = n(),
      
      chat_pct = chat_num / den,
      cla_pct  = cla_num / den,
      gem_pct  = gem_num / den,
      
      .groups = "drop"
    ) %>%
    mutate(
      variable = base,
      chat_nd  = paste0(chat_num, "/", den),
      cla_nd   = paste0(cla_num, "/", den),
      gem_nd   = paste0(gem_num, "/", den)
    ) %>%
    select(variable, human,
           chat_nd, chat_pct,
           cla_nd, cla_pct,
           gem_nd, gem_pct)
}




```

## Function to create heatmap

The `make_llm_agreement_heatmap()` function creates a paper-by-variable heatmap showing how many of the three language models matched the human-reviewed label for each observation. It can optionally relabel variables using label_list, reorder rows and columns by average agreement, and customize the plot title and subtitle. This figure is used to visualize where model agreement was high, where it was mixed, and which variables were most difficult for the models to classify consistently.

```{r, eval=FALSE}



make_llm_agreement_heatmap <- function(df,
                                       vars,
                                       paper_id_col = "filename",
                                       label_list = NULL,
                                       title = "LLM agreement heatmap",
                                       subtitle = "Each cell shows how many of 3 models matched the human label",
                                       low_to_high = TRUE,
                                       reorder = TRUE) {
  
  stopifnot(is.data.frame(df))
  stopifnot(is.character(vars))
  stopifnot(length(paper_id_col) == 1)
  
  if (!paper_id_col %in% names(df)) {
    stop("`paper_id_col` not found in `df`.")
  }
  
  needed_cols <- c(
    unlist(purrr::map(vars, ~ c(
      paste0(.x, "_human"),
      paste0(.x, "_ChatGPT"),
      paste0(.x, "_claude"),
      paste0(.x, "_gemini")
    ))),
    paper_id_col
  )
  
  missing_cols <- setdiff(needed_cols, names(df))
  if (length(missing_cols) > 0) {
    stop(
      "These required columns are missing from `df`: ",
      paste(missing_cols, collapse = ", ")
    )
  }
  
  heatmap_df <- purrr::map_dfr(vars, function(v) {
    human_col <- paste0(v, "_human")
    chat_col  <- paste0(v, "_ChatGPT")
    cla_col   <- paste0(v, "_claude")
    gem_col   <- paste0(v, "_gemini")
    
    pretty_label <- if (!is.null(label_list) && v %in% names(label_list)) {
      unname(label_list[[v]])
    } else {
      v
    }
    
    tibble::tibble(
      paper_id = stringr::word(as.character(df[[paper_id_col]]), 1) |>
        stringr::str_remove("\\.$"),
      variable = v,
      label = pretty_label,
      human = df[[human_col]],
      chat  = df[[chat_col]],
      cla   = df[[cla_col]],
      gem   = df[[gem_col]]
    ) |>
      dplyr::mutate(
        chat_match = dplyr::if_else(is.na(human) | is.na(chat), NA_integer_, as.integer(human == chat)),
        cla_match  = dplyr::if_else(is.na(human) | is.na(cla),  NA_integer_, as.integer(human == cla)),
        gem_match  = dplyr::if_else(is.na(human) | is.na(gem),  NA_integer_, as.integer(human == gem)),
        n_models_scorable = rowSums(!is.na(dplyr::pick(chat_match, cla_match, gem_match))),
        n_models_correct  = rowSums(dplyr::pick(chat_match, cla_match, gem_match), na.rm = TRUE),
        n_models_correct  = dplyr::if_else(n_models_scorable == 0, NA_integer_, n_models_correct)
      )
  })
  
  if (reorder) {
    row_summary <- heatmap_df |>
      dplyr::group_by(paper_id) |>
      dplyr::summarise(mean_models_correct = mean(n_models_correct, na.rm = TRUE), .groups = "drop")
    
    col_summary <- heatmap_df |>
      dplyr::group_by(label) |>
      dplyr::summarise(mean_models_correct = mean(n_models_correct, na.rm = TRUE), .groups = "drop")
    
    if (low_to_high) {
      row_order <- row_summary |>
        dplyr::arrange(mean_models_correct) |>
        dplyr::pull(paper_id)
      
      col_order <- col_summary |>
        dplyr::arrange(mean_models_correct) |>
        dplyr::pull(label)
    } else {
      row_order <- row_summary |>
        dplyr::arrange(dplyr::desc(mean_models_correct)) |>
        dplyr::pull(paper_id)
      
      col_order <- col_summary |>
        dplyr::arrange(dplyr::desc(mean_models_correct)) |>
        dplyr::pull(label)
    }
  } else {
    row_order <- df[[paper_id_col]] |>
      as.character() |>
      stringr::word(1) |>
      stringr::str_remove("\\.$") |>
      unique()
    
    col_order <- purrr::map_chr(vars, function(v) {
      if (!is.null(label_list) && v %in% names(label_list)) {
        unname(label_list[[v]])
      } else {
        v
      }
    })
  }
  
  heatmap_df <- heatmap_df |>
    dplyr::mutate(
      paper_id = factor(paper_id, levels = rev(row_order)),
      label = factor(label, levels = col_order),
      n_models_correct_f = factor(as.character(n_models_correct), levels = c("0", "1", "2", "3"))
    )
  
  # invisible layer to force all 4 legend boxes to appear, using geom_tile
  legend_df <- tibble::tibble(
    label = factor(rep(col_order[1], 4), levels = col_order),
    paper_id = factor(rep(rev(row_order)[1], 4), levels = levels(heatmap_df$paper_id)),
    n_models_correct_f = factor(c("0", "1", "2", "3"), levels = c("0", "1", "2", "3"))
  )
  
  ggplot2::ggplot(
    heatmap_df,
    ggplot2::aes(x = label, y = paper_id, fill = n_models_correct_f)
  ) +
    ggplot2::geom_tile(color = "white", linewidth = 0.25) +
    ggplot2::geom_tile(
      data = legend_df,
      ggplot2::aes(x = label, y = paper_id, fill = n_models_correct_f),
      inherit.aes = FALSE,
      alpha = 0,
      show.legend = TRUE
    ) +
    ggplot2::scale_fill_manual(
      values = c(
        "0" = "#b2182b",
        "1" = "#ef8a62",
        "2" = "#fddbc7",
        "3" = "#d1e5f0"
      ),
      breaks = c("0", "1", "2", "3"),
      drop = FALSE,
      name = "Models matching\nhuman"
    ) +
    ggplot2::guides(
      fill = ggplot2::guide_legend(
        override.aes = list(alpha = 1)
      )
    ) +
    ggplot2::labs(
      title = title,
      subtitle = subtitle,
      x = NULL,
      y = NULL
    ) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      panel.grid = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
      axis.text.y = ggplot2::element_text(size = 7)
    )
}


```

