---
title: "LLM data analysis: transplant"
format: html
---

::: {.Rcode title="Source code"}


The full R script is available at:

-   [`R/Rscript%20data%20analysis%20ATC%202025-11-16.R`](https://github.com/VagishHemmige/USRDS-PDF-analysis-project/blob/master/R/Rscript%20data%20analysis%20ATC%202025-11-16.R)


This R script file is itself reliant on the following helper files:

-   [`R/Analysis setup.R`](https://github.com/VagishHemmige/USRDS-PDF-analysis-project/blob/master/R/Analysis%20setup.R)
-   [`R/functions.R`](https://github.com/VagishHemmige/USRDS-PDF-analysis-project/blob/master/R/functions.R)
:::


# Overview

This page explains the R code used to generate the transplant-focused results for the ATC analysis. The script starts from a wide-format dataset containing human review results and outputs from multiple large language models (LLMs), then produces summary tables and agreement visualizations.

The overall workflow has four main parts. First, the dataset is restricted to papers that were classified by human review as both **new data analyses** and **transplant-related studies**. Second, the script creates summary tables showing runtime and variable-level agreement between each LLM and the human reference abstraction. Third, it builds descriptive breakdown tables that show how often each model matched the human classification for each response category. Finally, it generates heatmaps to visualize agreement patterns across papers and variables.

# Restricting the analytic dataset

The first step creates the analysis dataset used throughout the remainder of the script.

```{r eval=FALSE}
df_filt_atc <- df_wide %>%
  filter(
    is_this_paper_a_new_data_analysis_human == TRUE,
    is_this_a_transplant_related_study_human == TRUE
  )
```

This code filters the full wide-format dataset to include only papers that the human reviewer classified as both:

1. a new data analysis, and  
2. a transplant-related study.

This restriction is important because the downstream ATC results are intended to describe model performance within the subset of papers that were truly relevant to the transplant analysis. By applying the human-reviewed criteria first, the script ensures that agreement metrics are calculated on the final analytic sample rather than on the broader screening dataset.

# Table of model runtimes

The next section formats a simple table showing how long each LLM took to process the full dataset.

```{r eval=FALSE}
timing_gt<-timing_table %>%
  gt() %>%
  fmt_number(
    columns = `Time (minutes)`,
    decimals = 0
  )%>%
  tab_header(
    title = "Runtime of LLMs Used in the Analysis",
    subtitle = "Wall-clock time in minutes to process the full dataset"
  )

gt::gtsave(
  data = timing_gt,
  filename = "ATC 2026 abstract/timing_table_atc.png"
)
saveRDS(timing_gt, "ATC 2026 abstract/timing_table_atc.rds")
```

Here, the object `timing_table` is converted into a formatted `gt` table. The reported quantity is wall-clock time in minutes, which reflects the total elapsed time needed for each model to run across the full dataset. The table is then saved in two formats:

- a `.png` image for insertion into slides, posters, or webpages, and  
- an `.rds` object for later reuse in Quarto pages or other R scripts.

This pattern appears repeatedly in the script: create a formatted object, export an image version for presentation, and save an `.rds` version for reproducible reuse.

# Computing variable-level Cohen's kappa

The next section computes Cohen's kappa for each study variable and for each model, comparing model abstraction against the human reference standard.

```{r eval=FALSE}
kappa_results_atc <- purrr::map_dfr(human_vars, function(h) {
  base <- sub("_human$", "", h)

  tibble::tibble(
    variable = base,
    ChatGPT = compute_kappa(h, paste0(base, "_ChatGPT"), df_filt_atc),
    Claude  = compute_kappa(h, paste0(base, "_claude"),  df_filt_atc),
    Gemini  = compute_kappa(h, paste0(base, "_gemini"),  df_filt_atc)
  )%>%
    mutate(
    across(c(ChatGPT, Claude, Gemini), ~ ifelse(is.nan(.x), NA_real_, .x))
  )
})%>%
  filter(!variable %in% c(
    "which_pieces_of_data_were_abstracted_from_the_files",
    "which_tasks_would_not_be_covered",
    "which_type_of_transplant_study_is_this",
    "who_funded_the_study",
    "year",
    "journal",
    "title",
    "authors",
    "languages_used_SQL",
    "languages_used_Matlab",
    "languages_used_Julia"
  ))
```

This block loops over the vector `human_vars`, which contains the names of variables derived from human review. For each variable, the script removes the `_human` suffix to recover the shared base name, then uses that base name to identify the corresponding variables for ChatGPT, Claude, and Gemini.

The helper function `compute_kappa()` is then called three times per variable, once for each model. The result is a tidy table with one row per variable and one column per model.

After the kappas are computed, the script converts `NaN` values to missing values (`NA`). This is useful because some variables do not have enough variation, or may not have enough usable paired observations, to support estimation of Cohen's kappa. Treating these as missing makes them easier to handle in later formatting steps.

Finally, a small group of variables is excluded from the kappa table. These are mostly free-text metadata fields or rarely used categories that were not intended to appear in the final agreement summary.

# Formatting the kappa results as a table

The raw kappa results are then transformed into a publication-style table.

```{r eval=FALSE}
kappa_table_atc <- kappa_results_atc %>%
  add_label_for_gt() %>%
  gt(rowname_col = "label") %>%
  gt::cols_hide(columns = variable) %>%
  fmt_number(
    columns = c(ChatGPT, Claude, Gemini),
    decimals = 3
  ) %>%
  cols_align_decimal(
    columns = c(ChatGPT, Claude, Gemini)
  ) %>%
  tab_header(
    title = "Cohen’s κ: LLM vs Human Across All Variables",
    subtitle = "Dataset of 44 transplant-related papers"
  ) %>%
  gt::tab_stubhead(label = "Variable") %>%
  sub_missing(
    columns = c(ChatGPT, Claude, Gemini),
    missing_text = "-"
  )%>%
  gt::tab_row_group(
    label = "Programming languages",
    rows = grepl("^languages_", variable)
  ) %>%
  gt::tab_row_group(
    label = "Files used",
    rows = grepl("^files_", variable)
  )%>%
  gt::tab_row_group(
    label = "Components of USRDS used",
    rows = grepl("^component_", variable)
  ) %>%
  gt::tab_stub_indent(
    rows = grepl("^(languages_|files_|component_)", variable),
    indent = 4
  )%>%
  tab_footnote(
    footnote = "κ not estimable, usually due to insufficient variation or no usable paired observations.",
    locations = cells_body(columns = ChatGPT, rows = is.na(ChatGPT)),
    placement="right"
  ) %>%
  tab_footnote(
    footnote = "κ not estimable, usually due to insufficient variation or no usable paired observations.",
    locations = cells_body(columns = Claude, rows = is.na(Claude)),
    placement="right"
  ) %>%
  tab_footnote(
    footnote = "κ not estimable, usually due to insufficient variation or no usable paired observations.",
    locations = cells_body(columns = Gemini, rows = is.na(Gemini)),
    placement="right"
  )%>%
  data_color(
    columns = c(ChatGPT, Claude, Gemini),
    fn = kappa_fill
  )

kappa_table_atc
gtsave(
  data = kappa_table_atc,
  filename = "ATC 2026 abstract/kappa_table_atc.png"
)
saveRDS(kappa_table_atc, "ATC 2026 abstract/kappa_table_atc.rds")
```

This section takes the raw kappa results and makes them easier to interpret. The helper function `add_label_for_gt()` adds a human-readable label for each variable, allowing the final table to display descriptive names instead of internal variable names.

Several formatting choices improve readability:

- kappas are shown to three decimal places,
- numeric columns are aligned on the decimal point,
- missing kappas are displayed as `-` rather than `NA`,
- related variables are grouped into sections such as programming languages, files used, and USRDS components used, and
- indented rows visually distinguish subgroup items from the rest of the table.

The repeated footnotes explain why some kappa values are missing. This is an important interpretive detail, because missing kappa values usually reflect limited variation or insufficient paired observations rather than a coding error.

Finally, the call to `data_color()` applies the custom `kappa_fill()` function to color-code cells by agreement level. This makes it easier to scan the table and quickly identify variables with lower or higher agreement.

# Kappa legend

Because the kappa table is color-coded, the script also generates a companion legend.

```{r eval=FALSE}
kappa_legend_gt<-create_kappa_legend_gt()
gtsave(
  data = kappa_legend_gt,
  filename = "ATC 2026 abstract/kappa_table_legend_atc.png"
)
saveRDS(kappa_legend_gt, "ATC 2026 abstract/kappa_legend_gt.rds")
```

The helper function `create_kappa_legend_gt()` creates a small legend table that explains the meaning of the color scale used in the main kappa table. This is especially helpful for posters, slides, or webpages where readers may not immediately know how the shading maps to different levels of agreement.

# Building descriptive breakdown tables

In addition to kappa, the script generates a more descriptive table that shows the raw proportions of model responses within each human-coded category.

```{r eval=FALSE}
breakdown_table_atc <- map_dfr(human_vars, get_breakdown, df_filt = df_filt_atc)
```

This line applies the helper function `get_breakdown()` across all human-reviewed variables. Rather than summarizing agreement as a single kappa statistic, this step produces numerator/denominator counts and percentages for each model within each human category. These descriptive summaries complement the kappa table by showing the underlying distribution of matches.

The resulting dataset is then formatted as a `gt` table.

```{r eval=FALSE}
breakdown_gt_atc <- breakdown_table_atc %>%
  add_label_for_gt() %>%
  dplyr::select(-variable) %>%
  gt(
    groupname_col = "label",
    rowname_col = "human"
  ) %>%
  fmt_percent(
    columns = c(chat_pct, cla_pct, gem_pct),
    decimals = 1
  ) %>%
  cols_label(
    chat_nd = "Num/Den",
    chat_pct = "%",
    cla_nd = "Num/Den",
    cla_pct = "%",
    gem_nd = "Num/Den",
    gem_pct = "%"
  ) %>%
  tab_stubhead(label = md("**Human**")) %>%
  tab_spanner(
    label = md("**ChatGPT**"),
    columns = c(chat_nd, chat_pct)
  ) %>%
  tab_spanner(
    label = md("**Claude**"),
    columns = c(cla_nd, cla_pct)
  ) %>%
  tab_spanner(
    label = md("**Gemini**"),
    columns = c(gem_nd, gem_pct)
  ) %>%
  tab_options(
    row_group.as_column = TRUE,
    row_group.font.weight = "bold",
    table.font.size = px(13),
    heading.align = "left"
  )

breakdown_gt_atc
gtsave(
  data = breakdown_gt_atc,
  filename = "ATC 2026 abstract/breakdown_gt_atc.png"
)
saveRDS(breakdown_gt_atc, "ATC 2026 abstract/breakdown_gt_atc.rds")
```

This table is designed to answer a different question from the kappa table. Instead of summarizing agreement with one statistic per variable, it shows the raw counts and percentages for each model. The final layout places each model under its own column spanner, with paired columns for:

- numerator/denominator (`Num/Den`), and  
- percentage (`%`).

This is useful because a kappa value alone can sometimes hide important details about class imbalance or sparse categories. The breakdown table provides a more transparent view of how each model behaved for each response option.

# Heatmaps of agreement patterns

The last major section of the script produces heatmaps. These visualize agreement patterns across papers for different families of variables.

## Files used

```{r eval=FALSE}
file_vars_atc <- names(label_list)[str_detect(names(label_list), "^files_")]

files_res_atc <- make_llm_agreement_heatmap(
  df = df_filt_atc,
  vars = file_vars_atc,
  paper_id_col = "filename",
  label_list = label_list,
  title = "Files used heatmap",
  reorder = FALSE
)

files_res_atc
ggsave(
  plot = files_res_atc,
  filename = "ATC 2026 abstract/files_res_atc.png"
)
```

This block identifies variables whose names begin with `files_`, which represent specific USRDS files reported as used in each paper. Those variables are passed to `make_llm_agreement_heatmap()`, which generates a heatmap summarizing model-versus-human agreement by paper.

## Programming languages used

```{r eval=FALSE}
languages_vars_atc <- names(label_list)[str_detect(names(label_list), "^languages_")]

languages_res_atc <- make_llm_agreement_heatmap(
  df = df_filt_atc,
  vars = languages_vars_atc,
  paper_id_col = "filename",
  label_list = label_list,
  title = "Languages used heatmap",
  reorder = FALSE
)

languages_res_atc
ggsave(
  plot = languages_res_atc,
  filename = "ATC 2026 abstract/languages_res_atc.png"
)
```

This section repeats the same process for variables beginning with `languages_`. These heatmaps help show whether agreement differs across programming language variables, which may be especially sparse or inconsistently reported in the literature.

## Components of USRDS used

```{r eval=FALSE}
component_vars_atc <- names(label_list)[str_detect(names(label_list), "^component_")]

component_res_atc <- make_llm_agreement_heatmap(
  df = df_filt_atc,
  vars = component_vars_atc,
  paper_id_col = "filename",
  label_list = label_list,
  title = "Components used heatmap",
  reorder = FALSE
)

component_res_atc
ggsave(
  plot = component_res_atc,
  filename = "ATC 2026 abstract/component_res_atc.png"
)
```

Here the script focuses on variables describing which components of the USRDS dataset were used in each paper. Grouping these together allows the heatmap to show whether some parts of the registry were easier for models to identify than others.

## Other study features

```{r eval=FALSE}
other_vars_atc <- setdiff(
  setdiff(
    names(label_list),
    c(file_vars_atc, languages_vars_atc, component_vars_atc)
  ),
  "did_the_authors_explicitly_state_what_files_they_used"
)

other_res_atc <- make_llm_agreement_heatmap(
  df = df_filt_atc,
  vars = other_vars_atc,
  paper_id_col = "filename",
  label_list = label_list,
  title = "Other study features heatmap",
  reorder = FALSE
)

other_res_atc

ggsave(
  filename = "ATC 2026 abstract/other_res_atc.png",
  plot = other_res_atc,
)
```

The final heatmap brings together the remaining variables that do not fall into the prior groups. One variable is explicitly excluded to avoid redundancy. This produces a catch-all visualization for the rest of the study characteristics included in the abstraction schema.

Across all four heatmap sections, the helper function `make_llm_agreement_heatmap()` is doing the main work. Although its internal code is not shown here, its purpose is to transform variable-level agreement data into a paper-by-variable visual summary. This is useful because the tables describe average or category-level agreement, while the heatmaps reveal where disagreement clusters across individual papers.

# Why both tables and heatmaps were used

The script intentionally uses both tabular summaries and graphical summaries because they answer different interpretive questions.

The kappa table provides a compact measure of agreement for each variable and each model. The breakdown table adds transparency by showing the underlying numerators, denominators, and percentages within each human-coded category. The heatmaps then add a paper-level view, showing where agreement and disagreement are concentrated across the dataset.

Together, these outputs provide a more complete picture of model performance than any single summary metric alone.

# Outputs produced by the script

The script saves several reusable outputs for downstream use in Quarto pages, posters, and presentations:

- `timing_table_atc.png` and `timing_table_atc.rds`
- `kappa_table_atc.png` and `kappa_table_atc.rds`
- `kappa_table_legend_atc.png` and `kappa_legend_gt.rds`
- `breakdown_gt_atc.png` and `breakdown_gt_atc.rds`
- `files_res_atc.png`
- `languages_res_atc.png`
- `component_res_atc.png`
- `other_res_atc.png`

These saved objects make it possible to separate the computationally intensive analysis script from the reporting layer. In practice, the analysis can be run once, and the resulting images or `.rds` objects can then be loaded into Quarto pages without recomputing all tables and figures each time the site is rendered.
