---
title: "Manual data analysis"
format: html
---

::: {.Rcode title="Source code"}

The full R script is available at:

- [`R/Rscript letter analysis of manual review.R`](https://github.com/VagishHemmige/USRDS-PDF-analysis-project/blob/master/R/Rscript%20letter%20analysis%20of%20manual%20review.R)

This R script relies on the following helper files:

- [`R/Analysis setup.R`](https://github.com/VagishHemmige/USRDS-PDF-analysis-project/blob/master/R/Analysis%20setup.R)
- [`R/functions.R`](https://github.com/VagishHemmige/USRDS-PDF-analysis-project/blob/master/R/functions.R)

:::

This script prepares the manually reviewed dataset for descriptive analysis. It imports the abstraction spreadsheet, filters to confirmed records, creates derived variables from free-text fields, generates a study selection flowchart, and produces the tables used in the manual review results pages. The workflow is shown below in the same sequence used in the analysis script. 

## Import the manually reviewed data

The analysis begins by importing the reviewer spreadsheet from Excel. This file contains the manually abstracted information for each paper.

```{r, eval=FALSE}
Anagha_summer_project_form_2025_raw <- read_excel(
  "data-raw/Data abstraction Excel reviewer 1 and 2.xlsx",
  sheet = "Reviewer 2"
)
```

## Clean the manually reviewed dataset

The imported spreadsheet is standardized using `clean_names()`, and the analysis is restricted to records that were confirmed for inclusion. This creates the core cleaned dataset used in subsequent steps.

```{r, eval=FALSE}
Anagha_summer_project_form_clean <- Anagha_summer_project_form_2025_raw %>%
  clean_names() %>%
  filter(confirmed_by_vh == "Yes")
```

## Create the study selection flowchart

A flowchart is generated from the cleaned dataset to summarize how papers moved through the selection process. The diagram begins with the initial PubMed search results, then filters to new patient-level data analyses and splits the remaining papers into transplant-related and non-transplant-related studies.

```{r, eval=FALSE}
Anagha_summer_project_form_clean_fc <- Anagha_summer_project_form_clean %>%
  as_fc(label = "Initial papers from Pubmed search") %>%
  fc_filter(
    is_this_paper_a_new_data_analysis == "Yes",
    label = "New patient-level data analysis",
    show_exc = TRUE
  ) %>%
  fc_split(
    is_this_a_transplant_related_study,
    label = c("Not transplant-related", "Transplant-related")
  )
```

The resulting flowchart is then drawn and exported as a PNG image for use in the Quarto site.

```{r, eval=FALSE}
fc_draw(Anagha_summer_project_form_clean_fc) %>%
  fc_export("Letter analysis/flowchart.png")
```

## Create a table of included papers

The next step is to generate a formatted table listing the papers included in the manual analysis. This table is restricted to new patient-level data analyses and includes the main bibliographic details together with transplant-related status.

```{r, eval=FALSE}
paper_table_gt <- Anagha_summer_project_form_clean %>%
  filter(is_this_paper_a_new_data_analysis == "Yes") %>%
  select(
    authors,
    title,
    journal,
    year,
    transplant_related = is_this_a_transplant_related_study
  ) %>%
  distinct() %>%
  arrange(transplant_related, authors, desc(year), title) %>%
  gt() %>%
  cols_label(
    authors = "Authors",
    title = "Title",
    journal = "Journal",
    year = "Year",
    transplant_related = "Transplant related"
  ) %>%
  tab_header(
    title = "Papers Included in the Analysis"
  )
```

The formatted paper table is then saved for later reuse in the Quarto pages.

```{r, eval=FALSE}
paper_table_gt
saveRDS(paper_table_gt, file = "Letter analysis/paper_table_gt.rds")

gtsave(
  filename = "Letter analysis/paper_table_gt.html",
  data = paper_table_gt
)
```

## Create the final dataset for manual analysis

The confirmed dataset is then expanded into a more analysis-ready form. Several free-text fields are converted into binary indicator variables so that the contents of each paper can be summarized systematically. These include programming languages used, USRDS files used, and types of data components abstracted.

### Derive programming language indicators

```{r, eval=FALSE}
Anagha_summer_project_form_final <- Anagha_summer_project_form_clean_fc$data %>%
  mutate(languages_used_R = str_detect(what_programming_language_s_was_were_used, "R")) %>%
  mutate(languages_used_SAS = str_detect(what_programming_language_s_was_were_used, "SAS")) %>%
  mutate(languages_used_Stata = str_detect(what_programming_language_s_was_were_used, "STATA")) %>%
  mutate(languages_used_SPSS = str_detect(what_programming_language_s_was_were_used, "SPSS"))
```

### Derive file-use indicators

```{r, eval=FALSE}
Anagha_summer_project_form_final <- Anagha_summer_project_form_final %>%
  mutate(files_used_Core = str_detect(what_files_were_used, "Core")) %>%
  mutate(files_used_Institution = str_detect(what_files_were_used, "Institution")) %>%
  mutate(files_used_PhysicianSupplier = str_detect(what_files_were_used, "Physician/Supplier")) %>%
  mutate(files_used_Hospital = str_detect(what_files_were_used, "Hospital")) %>%
  mutate(files_used_Transplant = str_detect(what_files_were_used, "Transplant")) %>%
  mutate(files_used_PartD = str_detect(what_files_were_used, "Part D")) %>%
  mutate(files_used_CROWNWeb = str_detect(what_files_were_used, "CROWNWeb"))
```

### Derive abstracted data-component indicators

```{r, eval=FALSE}
Anagha_summer_project_form_final <- Anagha_summer_project_form_final %>%
  mutate(
    component_basicdemographics = str_detect(
      which_pieces_of_data_were_abstracted_from_the_files,
      "Basic demographics"
    )
  ) %>%
  mutate(
    component_icd9 = str_detect(
      which_pieces_of_data_were_abstracted_from_the_files,
      "ICD9 code"
    )
  ) %>%
  mutate(
    component_icd10 = str_detect(
      which_pieces_of_data_were_abstracted_from_the_files,
      "ICD10 code"
    )
  ) %>%
  mutate(
    component_cptcode = str_detect(
      which_pieces_of_data_were_abstracted_from_the_files,
      "CPT code"
    )
  ) %>%
  mutate(
    component_medicationD = str_detect(
      which_pieces_of_data_were_abstracted_from_the_files,
      "Medications"
    )
  ) %>%
  mutate(
    component_transplant = str_detect(
      which_pieces_of_data_were_abstracted_from_the_files,
      "Transplant-specific information"
    )
  ) %>%
  mutate(
    component_costs = str_detect(
      which_pieces_of_data_were_abstracted_from_the_files,
      "Costs"
    )
  ) %>%
  mutate(
    component_residential = str_detect(
      which_pieces_of_data_were_abstracted_from_the_files,
      "Residential location"
    )
  )
```

## Split the final dataset by transplant-related status

For some downstream analyses, the final dataset is divided into transplant-related and non-transplant-related subsets. This allows the descriptive summaries to be stratified by study type.

```{r, eval=FALSE}
Anagha_summer_project_form_split <- Anagha_summer_project_form_final %>%
  group_by(is_this_a_transplant_related_study) %>%
  group_split() %>%
  set_names(c("not_transplant", "transplant"))
```

## Create the overall summary table

The first descriptive table summarizes the full manually reviewed dataset after excluding identifying and free-text fields that are not intended for the summary output.

```{r, eval=FALSE}
letter_table1_gt <- Anagha_summer_project_form_final %>%
  select(
    -filepath,
    -timestamp,
    -authors,
    -title,
    -journal,
    -year,
    -what_programming_language_s_was_were_used,
    -what_files_were_used
  ) %>%
  tbl_summary() %>%
  as_gt()
```

The overall summary table is then saved for later rendering in the Quarto site.

```{r, eval=FALSE}
gtsave(
  filename = "Letter analysis/Letter analysis Table 1 overall raw.html",
  data = letter_table1_gt
)
saveRDS(
  letter_table1_gt,
  file = "Letter analysis/Letter analysis Table 1 overall raw.rds"
)
```

## Create the table stratified by transplant-related status

The next table summarizes the manually reviewed variables separately for transplant-related and non-transplant-related studies. Variable labels are applied using `label_list`, and related variables are grouped into sections for readability.

```{r, eval=FALSE}
letter_table2_gt <- Anagha_summer_project_form_final %>%
  select(
    is_this_a_transplant_related_study,
    did_the_authors_state_what_programming_language_they_used,
    languages_used_R,
    languages_used_SAS,
    languages_used_Stata,
    languages_used_SPSS,
    did_the_authors_explicitly_state_what_files_they_used,
    does_the_choice_of_files_codes_seem_appropriate_for_the_research_question,
    files_used_Core,
    files_used_Institution,
    files_used_PhysicianSupplier,
    files_used_Hospital,
    files_used_Transplant,
    files_used_PartD,
    files_used_CROWNWeb,
    component_basicdemographics,
    component_icd9,
    component_icd10,
    component_cptcode,
    component_medicationD,
    component_transplant,
    component_costs,
    component_residential,
    were_any_crosswalks_used,
    did_the_authors_share_the_code_they_used,
    was_the_study_funded,
    would_the_major_data_tasks_needed_for_this_study_be_covered_by_currently_proposed_usrds_r_package
  ) %>%
  tbl_summary(
    by = is_this_a_transplant_related_study,
    label = label_list,
    missing = "no"
  ) %>%
  modify_spanning_header(all_stat_cols() ~ "**Transplant-related study?**") %>%
  add_variable_group_header(
    header = "Data components abstracted",
    variables = c(
      component_basicdemographics,
      component_icd9,
      component_icd10,
      component_cptcode,
      component_medicationD,
      component_transplant,
      component_costs,
      component_residential
    )
  ) %>%
  add_variable_group_header(
    header = "Files used",
    variables = c(
      files_used_Core,
      files_used_Institution,
      files_used_PhysicianSupplier,
      files_used_Hospital,
      files_used_Transplant,
      files_used_PartD,
      files_used_CROWNWeb
    )
  ) %>%
  add_variable_group_header(
    header = "Programming language used",
    variables = c(
      languages_used_R,
      languages_used_SAS,
      languages_used_Stata,
      languages_used_SPSS
    )
  ) %>%
  modify_footnote_body(
    footnote = "Categories in this section are not mutually exclusive; percentages may not sum to 100%.",
    columns = label,
    rows = label %in% c(
      "Data components abstracted",
      "Files used",
      "Programming language used"
    )
  ) %>%
  as_gt()
```

This stratified table is also saved for reuse.

```{r, eval=FALSE}
saveRDS(
  letter_table2_gt,
  file = "Letter analysis/Letter analysis Table 2 stratified by transplant.rds"
)

gtsave(
  filename = "Letter analysis/Letter analysis Table 2 stratified by transplant.html",
  data = letter_table2_gt
)
```

## Create the table stratified by transplant study type

Among transplant-related studies, a further summary table is created using the more specific variable `which_type_of_transplant_study_is_this`. This allows the manual review results to be described in more detail within the transplant subset.

```{r, eval=FALSE}
letter_table3_gt <- Anagha_summer_project_form_split[["transplant"]] %>%
  select(
    which_type_of_transplant_study_is_this,
    did_the_authors_state_what_programming_language_they_used,
    languages_used_R,
    languages_used_SAS,
    languages_used_Stata,
    languages_used_SPSS,
    did_the_authors_explicitly_state_what_files_they_used,
    does_the_choice_of_files_codes_seem_appropriate_for_the_research_question,
    files_used_Core,
    files_used_Institution,
    files_used_PhysicianSupplier,
    files_used_Hospital,
    files_used_Transplant,
    files_used_PartD,
    files_used_CROWNWeb,
    component_basicdemographics,
    component_icd9,
    component_icd10,
    component_cptcode,
    component_medicationD,
    component_transplant,
    component_costs,
    component_residential,
    were_any_crosswalks_used,
    did_the_authors_share_the_code_they_used,
    was_the_study_funded,
    would_the_major_data_tasks_needed_for_this_study_be_covered_by_currently_proposed_usrds_r_package
  ) %>%
  tbl_summary(
    by = which_type_of_transplant_study_is_this,
    label = label_list,
    missing = "no"
  ) %>%
  modify_spanning_header(all_stat_cols() ~ "**Type of transplant study**") %>%
  add_variable_group_header(
    header = "Data components abstracted",
    variables = c(
      component_basicdemographics,
      component_icd9,
      component_icd10,
      component_cptcode,
      component_medicationD,
      component_transplant,
      component_costs,
      component_residential
    )
  ) %>%
  add_variable_group_header(
    header = "Files used",
    variables = c(
      files_used_Core,
      files_used_Institution,
      files_used_PhysicianSupplier,
      files_used_Hospital,
      files_used_Transplant,
      files_used_PartD,
      files_used_CROWNWeb
    )
  ) %>%
  add_variable_group_header(
    header = "Programming language used",
    variables = c(
      languages_used_R,
      languages_used_SAS,
      languages_used_Stata,
      languages_used_SPSS
    )
  ) %>%
  modify_footnote_body(
    footnote = "Categories in this section are not mutually exclusive; percentages may not sum to 100%.",
    columns = label,
    rows = label %in% c(
      "Data components abstracted",
      "Files used",
      "Programming language used"
    )
  ) %>%
  as_gt()
```

The final transplant-subtype table is saved for inclusion in the project outputs.

```{r, eval=FALSE}
saveRDS(
  letter_table3_gt,
  file = "Letter analysis/Letter analysis Table 3 stratified by transplant study type.rds"
)

gtsave(
  filename = "Letter analysis/Letter analysis Table 3 stratified by transplant study type.html",
  data = letter_table3_gt
)
```
