---
title: "Results: manual"
format: html
---



```{r, include=FALSE}


#This code prepares the .qmd environent for analysis


letter_table2_gt<-readRDS("../Letter analysis/Letter analysis Table 2 stratified by transplant.rds")


letter_table3_gt<-readRDS("../Letter analysis/Letter analysis Table 3 stratified by transplant study type.rds")



source("../R/Analysis setup.R")


```


## Characteristics of transplant-related versus non-transplant studies

This table compares manually reviewed papers according to whether the study was transplant-related. Overall, transplant-related and non-transplant studies were similar in many respects. In both groups, most papers stated the programming language used, and SAS was the most commonly reported language, with smaller proportions using R or Stata. Core USRDS files were used universally, and institution and Physician/Supplier files were used at similar rates across groups.

The largest differences were in the expected transplant-specific domains. Transplant-related studies were much more likely to use the transplant file and to abstract transplant-specific data elements, whereas non-transplant studies more often relied on hospital and Part D files. These findings suggest that most studies shared a common analytic foundation in standard USRDS files, but diverged in predictable ways depending on whether the research question required transplant-specific content.

Notably, explicit reporting of which USRDS files were used remained incomplete in both groups, despite most studies appearing to use files appropriate to the research question. Sharing of analytic code was rare across the manual review sample.

```{r, echo=FALSE}
letter_table2_gt
```

## Characteristics of transplant studies by study type

This table further stratifies transplant-related studies into post-transplant studies, pre-transplant access-to-care studies, and pre-transplant management or outcomes studies. Across all three subgroups, most papers reported the programming language used, suggesting relatively good reporting of analytic software among transplant-focused studies.

SAS was the dominant language across transplant study types, particularly in pre-transplant access-to-care studies, while R and Stata were used less often. Although the subgroup sizes are small, the overall pattern suggests that transplant studies varied more in clinical focus than in their basic computational approach. In other words, post-transplant and pre-transplant studies appeared to draw from a broadly similar analytic toolkit, with differences likely driven by the specific data elements needed for each research question.

```{r, echo=FALSE}
letter_table3_gt
```

## Summary

Taken together, these manually abstracted results suggest that published USRDS analyses rely on a fairly consistent computational backbone, centered on common source files and a limited set of statistical software platforms. The main differences between study groups arise from the clinical question being asked, particularly whether transplant-specific data are needed. At the same time, incomplete reporting of specific files used and the very low rate of code sharing highlight ongoing limitations in reproducibility.