---
title: "Setup"
format: html
---


The code in this script imports libraries, initializes constants, and performs other such preparatory work for the main analysis.

```{r, include=FALSE}

#Ensures pipe works
library(tidyverse)
library(gt)

```


::: {.Rcode title="Source code"}


The full R script is available at:

-   [`R/Analysis setup.R`](https://github.com/VagishHemmige/USRDS-PDF-analysis-project/blob/master/R/Analysis%20setup.R)

:::

## Libraries

The analysis relies on a set of open-source R packages for data management, cohort construction, visualization, and statistical modeling. These libraries were selected based on their stability, community support, and suitability for large-scale longitudinal health data. 


```{r eval=FALSE}

# Libraries necessary for the analysis

library(tidyverse)
library(readr)
library(readxl)
library(writexl)
library(janitor)
library(irr)
library(gtsummary)
library(gt)
library(officer)
library(tidyr)
library(stringr)
library(flowchart)
library(ellmer)
library(pdftools)
library(rstudioapi)
library(purrr)


```

The table below lists the required packages along with links to their primary documentation.

```{r, echo=FALSE}

libraries_table <- tibble::tibble(
  Category = c(
    "Core data handling",
    "Core data handling",
    "Core data handling",
    "Core data handling",
    "Core data handling",
    "Agreement and summaries",
    "Agreement and summaries",
    "Agreement and summaries",
    "Output and reporting",
    "Output and reporting",
    "Workflow and diagrams",
    "Workflow and diagrams",
    "PDF and LLM analysis",
    "PDF and LLM analysis",
    "PDF and LLM analysis"
  ),
  Package = c(
    "tidyverse",
    "readr",
    "readxl",
    "janitor",
    "writexl",
    "irr",
    "gtsummary",
    "gt",
    "officer",
    "writexl",
    "flowchart",
    "rstudioapi",
    "ellmer",
    "pdftools",
    "purrr"
  ),
  Documentation = c(
    "[tidyverse](https://www.tidyverse.org/)",
    "[readr](https://readr.tidyverse.org/)",
    "[readxl](https://readxl.tidyverse.org/)",
    "[janitor](https://sfirke.github.io/janitor/)",
    "[writexl](https://cran.r-project.org/package=writexl)",
    "[irr](https://cran.r-project.org/package=irr)",
    "[gtsummary](https://www.danieldsjoberg.com/gtsummary/)",
    "[gt](https://gt.rstudio.com/)",
    "[officer](https://davidgohel.github.io/officer/)",
    "[writexl](https://cran.r-project.org/package=writexl)",
    "[flowchart](https://cran.r-project.org/package=flowchart)",
    "[rstudioapi](https://rstudio.github.io/rstudioapi/)",
    "[ellmer](https://ellmer.tidyverse.org/)",
    "[pdftools](https://cran.r-project.org/package=pdftools)",
    "[purrr](https://purrr.tidyverse.org/)"
  )
)

libraries_table |>
  gt::gt() |>
  gt::fmt_markdown(columns = "Documentation") |>
  gt::cols_label(
    Category = "Category",
    Package = "R package",
    Documentation = "Documentation"
  ) |>
  gt::tab_header(
    title = "Required R Packages",
    subtitle = "Libraries used for PDF processing, structured LLM extraction, data cleaning, agreement analysis, and reporting"
  ) |>
  gt::tab_style(
    style = gt::cell_text(font = "mono"),
    locations = gt::cells_body(columns = Package)
  )

```


## Label dictionary and analysis objects

The PDF analysis uses a small set of shared objects defined in `R/Analysis setup.R`. These help standardize variable labels across tables and figures and store supporting information used in the results pages.

### Variable label dictionary

The label_list object maps internal variable names used in the analysis code to reader-friendly labels used in tables and summaries. This allows the same underlying variables to be displayed consistently across outputs without manually renaming them in each table.

```{r}



label_list <- list(
  did_the_authors_state_what_programming_language_they_used = "Programming language stated",
  did_the_authors_explicitly_state_what_files_they_used = "USRDS files explicitly stated",
  does_the_choice_of_files_codes_seem_appropriate_for_the_research_question = "Files/codes appropriate for research question",
  did_the_authors_share_the_code_they_used = "Authors shared code",
  was_the_study_funded = "Study funded",
  would_the_major_data_tasks_needed_for_this_study_be_covered_by_currently_proposed_usrds_r_package = "Covered by proposed usRds package",
  were_any_crosswalks_used = "Any crosswalks used",
  languages_used_R = "R used",
  languages_used_SAS = "SAS used",
  languages_used_Stata = "Stata used",
  languages_used_SPSS = "SPSS used",
  languages_used_Python = "Python used",
  
  files_used_Core = "Core file used",
  files_used_Institution = "Institution file used",
  files_used_PhysicianSupplier = "Physician/Supplier file used",
  files_used_Hospital = "Hospital file used",
  files_used_Transplant = "Transplant file used",
  files_used_PartD= "Part D file used",
  files_used_CROWNWeb = "CROWNWeb file used",
  
  component_basicdemographics = "Basic demographics abstracted",
  component_icd9 = "ICD-9 codes abstracted",
  component_icd10 = "ICD-10 codes abstracted",
  component_cptcode = "CPT codes abstracted",
  component_medicationD = "Part D medication data abstracted",
  component_transplant = "Transplant-specific data abstracted",
  component_costs = "Cost data abstracted",
  component_residential = "Residential location abstracted",
  
  is_this_a_transplant_related_study = "Transplant-related study",
  is_this_paper_a_new_data_analysis = "New patient-level data analysis"
)



```

For clarity, the label dictionary is summarized below along with brief explanations of their role in the analysis.



```{r, echo=FALSE}

label_table <- tibble::tibble(
  Variable = names(label_list),
  Label = unname(unlist(label_list))
)

label_table |>
  gt::gt() |>
  gt::tab_header(
    title = "Variable label dictionary",
    subtitle = "Internal variable names mapped to display labels used in tables and figures"
  ) |>
  gt::tab_style(
    style = gt::cell_text(font = "mono"),
    locations = gt::cells_body(columns = Variable)
  )

```




### Model runtime table

The timing_table object stores the approximate runtime required to process the full dataset with each model. This provides context for the practical tradeoff between speed and extraction performance.

```{r, eval=FALSE}

#Table of times to run the analysis
timing_table <- tibble(
  `Provider (Company)` = c(
    "ChatGPT (OpenAI)",
    "Claude (Anthropic)",
    "Gemini (Google)"
  ),
  Model = c(
    "Gpt 5.1",
    "Claude Sonnet 4.5",
    "Gemini 2.5 Flash"
  ),
  `Time (minutes)` = c(
    165,
    62,
    672
  )
)


```


