---
title: "Setup"
format: html
---


The code in this script imports libraries, initializes constants, and performs other such preparatory work for the main analysis.

```{r, include=FALSE}

#Ensures pipe works
library(tidyverse)
library(gt)

```


::: {.Rcode title="Source code"}


The full R script is available at:

-   [`R/setup.R`](https://github.com/VagishHemmige/HOPE-CDC-analysis-2026/blob/master/R/setup.R)

:::

## Libraries

The analysis relies on a set of open-source R packages for data management, cohort construction, visualization, and statistical modeling. These libraries were selected based on their stability, community support, and suitability for large-scale longitudinal health data. 


```{r eval=FALSE}

# Libraries necessary for the analysis

# Data management
library(tidyverse)
library(stringi)
library(glue)

#Core library for the SRTR
library(sRtr)

#Data import libraries
library(readxl)
library(readr)
library(haven)
library(mapboxapi)
library(CDCAtlas)

#Geographic libraries
library(sf)
library(tigris)
library(tidycensus)
library(tmap)
library(lutz)

#Table libraries
library(flextable)
library(gt)

#Plotting
library(colorspace)
library(ggtext)
library(extrafont)
library(janitor)
library(ggrepel)
library(writexl)
library(plotly)
library(htmlwidgets)
library(patchwork)
library(ggbeeswarm)
library(ggpubr)
library(cowplot)
library(scales)


```

The table below lists the required packages along with links to their primary documentation.

```{r, echo=FALSE}


libraries_table <- tibble::tibble(
  Category = c(
    "Data management",
    "Data management",
    "Data management",
    "SRTR-specific",
    "Data import",
    "Data import",
    "Data import",
    "Data import",
    "Data import",
    "Geographic",
    "Geographic",
    "Geographic",
    "Geographic",
    "Geographic",
    "Tables",
    "Tables",
    "Plotting",
    "Plotting",
    "Plotting",
    "Plotting",
    "Plotting",
    "Plotting",
    "Plotting",
    "Plotting",
    "Plotting",
    "Plotting",
    "Plotting",
    "Plotting",
    "Plotting"
  ),
  Package = c("tidyverse", 
              "stringi", 
              "glue",
              "sRtr",
              "readxl", 
              "readr", 
              "haven", 
              "mapboxapi", 
              "CDCAtlas",
              "sf", 
              "tigris", 
              "tidycensus", 
              "tmap", 
              "lutz",
              "flextable", 
              "gt",
              "colorspace", 
              "ggtext", 
              "extrafont", 
              "janitor", 
              "ggrepel", 
              "writexl",
              "plotly", 
              "htmlwidgets", 
              "patchwork", 
              "ggbeeswarm", 
              "ggpubr", 
              "cowplot", 
              "scales"
  )
  ,
Documentation = c(
    "[tidyverse](https://www.tidyverse.org/)",
    "[stringi](https://stringi.gagolewski.com/)",
    "[glue](https://glue.tidyverse.org/)",
    "[sRtr GitHub](https://github.com/vagishhemmige/sRtr)",

    "[readxl](https://readxl.tidyverse.org/)",
    "[readr](https://readr.tidyverse.org/)",
    "[haven](https://haven.tidyverse.org/)",
    "[mapboxapi](https://walker-data.com/mapboxapi/)",
    "[CDCAtlas GitHub](https://github.com/vagishhemmige/CDCAtlas)",

    "[sf](https://r-spatial.github.io/sf/)",
    "[tigris](https://walker-data.com/tigris/)",
    "[tidycensus](https://walker-data.com/tidycensus/)",
    "[tmap](https://r-tmap.github.io/tmap/)",
    "[lutz](https://cran.r-project.org/package=lutz)",

    "[flextable](https://davidgohel.github.io/flextable/)",
    "[gt](https://gt.rstudio.com/)",

    "[colorspace](https://colorspace.r-forge.r-project.org/)",
    "[ggtext](https://wilkelab.org/ggtext/)",
    "[extrafont](https://cran.r-project.org/package=extrafont)",
    "[janitor](https://sfirke.github.io/janitor/)",
    "[ggrepel](https://ggrepel.slowkow.com/)",
    "[writexl](https://cran.r-project.org/package=writexl)",

    "[plotly](https://plotly.com/r/)",
    "[htmlwidgets](https://www.htmlwidgets.org/)",
    "[patchwork](https://patchwork.data-imaginist.com/)",
    "[ggbeeswarm](https://cran.r-project.org/package=ggbeeswarm)",
    "[ggpubr](https://rpkgs.datanovia.com/ggpubr/)",
    "[cowplot](https://wilkelab.org/cowplot/)",
    "[scales](https://scales.r-lib.org/)"
  )
  )

libraries_table |>
  gt::gt() |>
  gt::fmt_markdown(columns = "Documentation") |>
  gt::cols_label(
    Category = "Category",
    Package = "R package",
    Documentation = "Documentation"
  ) |>
  gt::tab_header(
    title = "Required R Packages",
    subtitle = "Libraries used for data processing, visualization, and statistical modeling"
  )|>
  gt::tab_style(
  style = gt::cell_text(font = "mono"),
  locations = gt::cells_body(columns = Package)
)



```


## Constants

A number of analysis parameters are reused across cohort construction, matching, and cost modeling steps. 

These constants are defined in the `R/setup.R` file.

```{r}


organ_list<-c("liver", "kidney", "heart", "lung", "pancreas")
year_list<-c("2017", "2022")
distance_list=c("50mile", "60min", "100mile", "200mile")

#List states which are continental
continental_fips <- c(
  "01", "04", "05", "06", "08", "09", "10", "11", "12", "13", "16",
  "17", "18", "19", "20", "21", "22", "23", "24", "25", "26",
  "27", "28", "29", "30", "31", "32", "33", "34", "35", "36",
  "37", "38", "39", "40", "41", "42", "44", "45", "46", "47",
  "48", "49", "50", "51", "53", "54", "55", "56"
)

#For isochrone analysis, we need to account for time zones in order to calculate driving time at 7 AM on the specific date

#List US continental time zones

timezones<-c("America/New_York","America/Chicago","America/Denver","America/Los_Angeles","America/Phoenix",
             "America/Indiana/Indianapolis","America/Kentucky/Louisville","America/Detroit")

reference_date<-list()

# Specific date to anchor the time: First Monday in March of both years
reference_date[["2022"]] <- as.Date("2022-03-07")
reference_date[["2017"]] <- as.Date("2017-03-06")

```

For clarity, these constants are summarized below along with brief explanations of their role in the analysis.

```{r, echo=FALSE}

format_constant <- function(x) {
  if (inherits(x, "Date")) {
    format(x, "%Y-%m-%d")
    
  } else if (is.list(x) && all(sapply(x, inherits, "Date"))) {
    paste(
      paste(names(x), format(as.Date(unlist(x), origin = "1970-01-01"), "%Y-%m-%d"), sep = ": "),
      collapse = " | "
    )
    
  } else if (length(x) > 1) {
    paste(x, collapse = ", ")
    
  } else {
    as.character(x)
  }
}

constants_table <- tibble::tibble(
  Constant = c(
    "organ_list",
    "year_list",
    "distance_list",
    "continental_fips",
    "timezones",
    "reference_date"
  ),
  Value = c(
    format_constant(organ_list),
    format_constant(year_list),
    format_constant(distance_list),
    format_constant(continental_fips),
    format_constant(timezones),
    format_constant(reference_date)
  ),
  Explanation = c(
    "Organs included in transplant center access analysis",
    "Calendar years evaluated for geographic access comparison",
    "Distance or drive-time buffers used to define transplant center catchment areas",
    "FIPS codes representing continental U.S. states included in analysis",
    "Time zones used to convert local 7 AM departure times to UTC for isochrone calculations",
    "Anchor dates (first Monday in March) used for time-zone–adjusted isochrone generation"
  )
)

constants_table |>
  gt::gt() |>
  gt::tab_header(
    title = "Analysis Constants",
    subtitle = "Geographic buffer definitions and time-zone parameters"
  )


```


